---
title: "Fraud Detection on Bank Payments"
subtitle: "Download model artifact to deploy using Docker"
author: "domjoseph2000@gmail.com  "
---

This code downloads the .rds artifact from latest model in production to be used for deployment as a REST Api using Plumber and Docker.

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning=FALSE, message=FALSE}
library(mlflow)
```
Run the below lines in command line:
`conda activate mlflow-env`

`mlflow server --host 127.0.0.1 --port 5000 --backend-store-uri "file:///A:/Dominic/My Projects/Fraud Detection/scripts/mlruns" --default-artifact-root "file:///A:/Dominic/My Projects/Fraud Detection/scripts/mlruns"`

After running, Set the MLflow Trackig URI
```{r}
mlflow_set_tracking_uri("http://127.0.0.1:5000")
```

Fetch the latest versions for the registered model

```{r, warning=FALSE, message=FALSE}
# This function returns the latest versions for the registered model
versions <- mlflow_get_latest_versions("Production")
```


# Find the specific version

```{r}
# Find the specific version that has the "prod" alias
run_id_to_download = versions[[1]]$run_id
local_path <- mlflow_download_artifacts(
  run_id = run_id_to_download, 
  path = "modelobjects/fraud_model.rds" # artifact path of rds object
)

```

Now lets copy the data to `data` folder.

```{r, warning=FALSE}
setwd("../")
print(getwd())
destination_path = paste0(getwd(),"/data")
print(paste0("Copying file to ", destination_path))

```
```{r}
file.copy(from = trimws(local_path), to = destination_path, recursive = TRUE)
```