---
title: "Fraud Detection on Bank Payments"
subtitle: "Model Selection and Performance Analysis"
author: "domjoseph2000@gmail.com  "
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, warning=FALSE, message=FALSE}
source("../packages.R")
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(caret)
library(rsample)
#library(rpart)  # for decision tree
#library(rpart.plot)
#library(rattle) # for plotting decision tree
#library(e1071)  # For Naive Bayes classifier
library(naivebayes)
library(randomForest)
library(pROC)
library(mlflow)
library(carrier)
```


## **1. Import Data For Modelling**

```{r}
#transactions = read.csv("../data/transformed_data.csv",sep =";")
transactions <- readRDS("../data/transactions_intermediate.rds")
head(transactions)

```
Now lets look at the structure of th imported data.

```{r}
str(transactions)
```

## **2. Fitting a classification model**

### 2.1 sampling data

Create test and train data.
```{r}
set.seed(123)
split_strat <- initial_split(transactions, prop = 0.8,
                             strata = 'fraud')
train <- training(split_strat)
test  <- testing(split_strat)
#rm(transactions)
```

### 2.2 Mlflow experiment

Create an mlflow experiment and start a run

First, you need to specify the URI scheme (file:///) for the server and backend store. In a terminal, start the server pointing to your existing backend. Create a directory (for example `C:/Users/username/mlflow-backend` if you have to):

Run the below lines in command line:
```
conda activate mlflow-env

mlflow server --host 127.0.0.1 --port 5000 --backend-store-uri "file:///A:/Dominic/My Projects/Fraud Detection/mlruns" --default-artifact-root "file:///A:/Dominic/My Projects/Fraud Detection/mlruns"
```

```{r}
# --- 2. Set the MLflow tracking server ---
mlflow_set_tracking_uri("http://127.0.0.1:5000")
```

Then create and set the experiment as active. Un-comment 1st line if performing a run the first time.
```{r}
#mlflow_create_experiment("Fraud-Detection-MLmodels")
exp_location = "file:///A:/Dominic/My Projects/Fraud Detection/mlruns/artifacts"
expid = mlflow_set_experiment("Fraud-Detection-MLmodels", artifact_location = exp_location)
```

### 2.3 Train and Log Models

#### logistic regression model

Use train data to perform logistic regression.

```{r}
with(mlflow_start_run()->active_run, {
  
  # Set the run name using a tag right at the start 🏷️
  mlflow_set_tag("mlflow.runName", "LogisticRegression-base")
  # Get the details of the current active run
  current_run_id <- active_run$run_uuid
  run_folder = paste0("/", expid, "/", current_run_id)
  
  # Log parameters
  mlflow_log_param("model_type", "Logistic Regression (glm)")
  mlflow_log_param("parameters", "{family : binomial, link : logit}")

  model_glm <- glm(fraud ~ log_amount + gender + mod_age + mod_category, data = train, family = binomial(link = 'logit'))
  print(summary(model_glm))

  # --- Log Metrics ---
  # Evaluate the model on the test set
  predictions <- predict(model_glm, newdata = test, type = "response")
  
  # Calculate AUC (Area Under the Curve)
  roc_obj <- roc(response = test$fraud, predictor = as.vector(predictions))
  auc_metric <- as.numeric(auc(roc_obj))
  
  # Log the AUC metric to MLflow
  mlflow_log_metric("auc", auc_metric)
  message(paste("AUC:", round(auc_metric, 4)))
  
  # Extract confusion metrics
  response_binary = (predictions >= 0.4)*1
  cm = confusionMatrix(as.factor(response_binary), as.factor(test$fraud), positive = '1')
  sensitivity <- cm$byClass['Sensitivity']
  Precision <- cm$byClass['Pos Pred Value'] 
  f1_score <- cm$byClass['F1']
  
  # Log the metrics
  mlflow_log_metric("sensitivity", sensitivity)
  mlflow_log_metric("precision", Precision)
  mlflow_log_metric("f1_score", f1_score)
  
  message(paste("F1 Score:", round(f1_score, 4)))
  
  # Save the confusion matrix to a text file
  writeLines(capture.output(cm), "confusion_matrix.txt")
  
  # Log the confusion matrix as an artifact
  mlflow_log_artifact("confusion_matrix.txt",
                      artifact_path = "artifacts",
                      run_id = current_run_id
                      )
  # --- Create and Save the Plot to a File ---
  png(filename = 'ROC.png')
  plot(1-roc_obj$specificities, roc_obj$sensitivities, main = paste0("AUC = ", round(auc(roc_obj), 3)), xlab = 'false positive rate', ylab = 'True positive rate')
  abline(a = 0, b = 1, lty = 2, col = "gray") 
  dev.off()
  mlflow_log_artifact('ROC.png', artifact_path = "plots")
  
  # --- Log the Model ---
  
  # 1. Create a "predictor" function using carrier::crate() 📦
  # This wraps your model's predict method in a standard format.
  predictor <- carrier::crate(function(new_data) {
    stats::predict(!!model_glm, newdata = new_data, type = "response")
  },model = model_glm)
  
  # 2. Log the generic 'predictor' object instead of the raw glm model
  mlflow_log_model(
    predictor,
    artifact_path = "modelobjects"
    )
  # Save locally
  saveRDS(model_glm, "fraud_model.rds")
  # Log the RDS file as an artifact
  mlflow_log_artifact("fraud_model.rds", artifact_path = "modelobjects")
  
  message("Successfully logged parameters, metrics, and model to MLflow.")
  
})
```

#### Fit Naive Bayes model:

```{r}
with(mlflow_start_run()->active_run, {
  
  # Set the run name using a tag right at the start 🏷️
  mlflow_set_tag("mlflow.runName", "Naive Bayes-base")
  
  # Log parameters
  mlflow_log_param("model_type", "Naive Bayes")
  mlflow_log_param("parameters", "{laplace : 10, kernel : TRUE}")

  model_nb <- naive_bayes(factor(fraud) ~ log_amount + gender + mod_age + mod_category, data = train, usekernel = T, laplace = 10)
  print(summary(model_nb))

  # --- Log Metrics ---
  # Evaluate the model on the test set
  predictions <- stats::predict(model_nb, newdata = test, type = "prob")[,'1']
  
  # Calculate AUC (Area Under the Curve)
  roc_obj   <- roc(response = test$fraud, predictor = as.vector(predictions))
  auc_metric <- as.numeric(auc(roc_obj))
  
  # Log the AUC metric to MLflow
  mlflow_log_metric("auc", auc_metric)
  message(paste("AUC:", round(auc_metric, 4)))
  
  # Extract confusion metrics
  response_binary = (predictions >= 0.5)*1
  cm = confusionMatrix(as.factor(response_binary), as.factor(test$fraud), positive = '1')
  sensitivity <- cm$byClass['Sensitivity'] # Note: Sensitivity and Recall are the same
  Precision <- cm$byClass['Pos Pred Value'] 
  f1_score <- cm$byClass['F1']
  
  # Log the metrics
  mlflow_log_metric("sensitivity", sensitivity)
  mlflow_log_metric("precision", Precision)
  mlflow_log_metric("f1_score", f1_score)
  
  message(paste("F1 Score:", round(f1_score, 4)))
  
  # Save the confusion matrix to a text file
  writeLines(capture.output(cm), "confusion_matrix.txt")
  
  # Log the confusion matrix as an artifact
  mlflow_log_artifact("confusion_matrix.txt",
                      artifact_path = "artifacts"
                      )
  
   # --- Create and Save the Plot to a File ---
  png(filename = 'ROC.png')
  plot(1-roc_obj$specificities, roc_obj$sensitivities, main = paste0("AUC = ", round(auc(roc_obj), 3)), xlab = 'false positive rate', ylab = 'True positive rate')
  abline(a = 0, b = 1, lty = 2, col = "gray") 
  dev.off()
  mlflow_log_artifact('ROC.png', artifact_path = "plots")
  
  # --- Log the Model ---
  
  # 1. Create a "predictor" function using carrier::crate() 📦
  # This wraps your model's predict method in a standard format.
  predictor <- carrier::crate(function(new_data) {
    new_data$log_amount   <- as.numeric(new_data$log_amount)
    new_data$gender       <- as.character(new_data$gender)
    new_data$mod_age      <- as.character(new_data$mod_age)
    new_data$mod_category <- as.character(new_data$mod_category)
    
    stats::predict(!!model_nb, newdata = new_data, type = "prob")[,'1']
  },model = model_nb)
  
  # 2. Log the generic 'predictor' object instead of the raw glm model
  mlflow_log_model(
    predictor,
    artifact_path = "modelobjects"
    )
  # Save locally
  saveRDS(model_nb, "fraud_model.rds")
  # Log the RDS file as an artifact
  mlflow_log_artifact("fraud_model.rds", artifact_path = "modelobjects")

  message("Successfully logged parameters, metrics, and model to MLflow.")
  
})
```
#### Fit Random Forest model

Use train data to perform Random Forest.

```{r}
# --- 4. Train and Log Random Forest ---
with(mlflow_start_run(), {
  
  # Set the run name using a tag right at the start 🏷️
  mlflow_set_tag("mlflow.runName", "RandomForest-base")
  mlflow_log_param("model_type", "RandomForest")
  mlflow_log_param("parameters", "{ntree : 500, mtry : 2}")

  # Train model
  model_rf <- randomForest::randomForest(
    as.factor(fraud) ~ log_amount + gender + mod_age + mod_category, 
    data = train, 
    ntree = 500, # Number of trees in the forest
    mtry = 2 # Number of variables sampled at each split
  )
  print(summary(model_rf))
  # --- Log Metrics ---
  # Evaluate the model on the test set
  predictions <- predict(model_rf, newdata = test, type = "prob")[,'1']
  
  # Calculate AUC (Area Under the Curve)
  roc_obj   <- roc(response = test$fraud, predictor = as.vector(predictions))
  auc_metric <- as.numeric(auc(roc_obj))
  
  # Log the AUC metric to MLflow
  mlflow_log_metric("auc", auc_metric)
  message(paste("AUC:", round(auc_metric, 4)))
  
  # Extract confusion metrics
  response_binary = (predictions >= 0.5)*1
  cm = confusionMatrix(as.factor(response_binary), as.factor(test$fraud), positive = '1')
  sensitivity <- cm$byClass['Sensitivity'] # Note: Sensitivity and Recall are the same
  Precision <- cm$byClass['Pos Pred Value'] 
  f1_score <- cm$byClass['F1']
  
  # Log the metrics
  mlflow_log_metric("sensitivity", sensitivity)
  mlflow_log_metric("precision", Precision)
  mlflow_log_metric("f1_score", f1_score)
  
  message(paste("F1 Score:", round(f1_score, 4)))
  
  # Save the confusion matrix to a text file
  writeLines(capture.output(cm), "confusion_matrix.txt")
  
  # Log the confusion matrix as an artifact
  mlflow_log_artifact("confusion_matrix.txt",
                      artifact_path = "artifacts"
                      )
  
  # --- Create and Save the Plot to a File ---
  png(filename = 'ROC.png')
  plot(1-roc_obj$specificities, roc_obj$sensitivities, main = paste0("AUC = ", round(auc(roc_obj), 3)), xlab = 'false positive rate', ylab = 'True positive rate')
  abline(a = 0, b = 1, lty = 2, col = "gray") 
  dev.off()
  mlflow_log_artifact('ROC.png', artifact_path = "plots")
  
  # Create a crate for the model
  predictor <- crate(function(new_data) {
    stats::predict(!!model_rf, newdata = new_data, type = "prob")[,'1']
  },model = model_rf)
  # 2. Log the generic 'predictor' object instead of the raw glm model
  mlflow_log_model(
    predictor,
    artifact_path = "modelobjects"
    )
  
    # Save locally
  saveRDS(model_rf, "fraud_model.rds")
  # Log the RDS file as an artifact
  mlflow_log_artifact("fraud_model.rds", artifact_path = "modelobjects")
  
  message("Successfully logged parameters, metrics, and model to MLflow.")
})
```

#### Fit SVM model

Use train data to perform SVM.
```{r,warning=FALSE, message=FALSE}
# to be updated

```

## **6. Conclusion**

```{r}
# Search for all runs within that experiment using its ID
runs_df <- mlflow_search_runs(experiment_id = expid)

# Filter the metrics df
merics_df = runs_df %>% 
  select(run_name, metrics) %>%
  unnest(metrics) %>%
  bind_rows(.) %>%
  as.data.frame(.) %>%
  pivot_wider(names_from = 'key', 
              values_from = 'value',
              id_cols = c(run_name,run_id),
              values_fn = max
                )
# Print the resulting table to the console
print(merics_df)
```

Out of the models, each model has its own strength in each of the metrics considered. 
Logistic Regression is observed to perform well on all 3 metrics, displaying an AUC value of 96% and a sensitivity of 62%,combined with an F1 score of 70%. Random forest had a slightly higher F1 score (72%), combined with high precision but lower AUC and sensitivity. While Naive Bayes has the highest AUC score at 97.7%, and also a sensitivity of 72%, which is highest among all 3,But F1 score of 56% is lowest among the three. 



